% SPDX-FileCopyrightText: 2024 Maxim Urschumzew <mxmurw@determi.io>
% SPDX-FileCopyrightText: 2024 Miëtek Bak
%
% SPDX-License-Identifier: MIT

\documentclass{scrartcl}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bm}
\usepackage{enumitem}
\usepackage{float}
\usepackage{graphicx}
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{tikz}
\usepackage[hyphens]{url}
\usepackage{xcolor}

% hide quotation marks in bibliography
\usepackage{csquotes}
\DeclareQuoteStyle{none}{}{}{}{}
\setquotestyle{none}

% use biblatex for bibliography
\usepackage[backend=biber,
            doi=true,
            maxnames=10,
            natbib=true,
            url=true]{biblatex}
\addbibresource{main.bib}
\DeclareFieldFormat{url}{\url{#1}}
\renewcommand{\intitlepunct}{~}

% use microtype for advanced typography
\usepackage[activate={true,nocompatibility},
            final,
            tracking=true,
            kerning=true,
            spacing=true,
            factor=1100,
            stretch=10,
            shrink=10]{microtype}
\microtypecontext{spacing=nonfrench}
\DeclareMicrotypeSet*[tracking]{ttfonts}{family=tt*}
\SetTracking{encoding={*}, family=tt*}{-10}
\SetTracking{encoding={*}, shape=sc}{0}

% use hyperref last to avoid problems
\usepackage{hyperref}
\definecolor{urlblue}{rgb}{0,0.2,0.8}
\hypersetup{
  allcolors=urlblue,
  bookmarksnumbered=false,
  bookmarksopen=false,
  breaklinks=true,
  colorlinks,
  pdfborder={0 0 0},
  pdfborderstyle={/S/U/W 1}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}[definition]{Example}
\theoremstyle{plain}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{corollary}[definition]{Corollary}
\newtheorem{conjecture}[definition]{Conjecture}

\renewcommand{\circle}{\mathchoice%
  {\tikz{\draw (0,0) circle [radius=0.1cm];}}%
  {\tikz{\draw (0,0) circle [radius=0.1cm];}}%
  {\tikz{\draw (0,0) circle [radius=0.075cm];}}%
  {\tikz{\draw (0,0) circle [radius=0.05cm];}}}

\renewcommand{\triangle}{\mathchoice%
  {\tikz{\draw (0,0) -- (0.2cm,0) -- (0.1cm,0.2cm) -- cycle;}}%
  {\tikz{\draw (0,0) -- (0.2cm,0) -- (0.1cm,0.2cm) -- cycle;}}%
  {\tikz{\draw (0,0) -- (0.15cm,0) -- (0.075cm,0.15cm) -- cycle;}}%
  {\tikz{\draw (0,0) -- (0.10cm,0) -- (0.05cm,0.10cm) -- cycle;}}}

\renewcommand{\square}{\mathchoice%
  {\tikz{\draw (0,0) rectangle (0.2cm,0.2cm);}}%
  {\tikz{\draw (0,0) rectangle (0.2cm,0.2cm);}}%
  {\tikz{\draw (0,0) rectangle (0.15cm,0.15cm);}}%
  {\tikz{\draw (0,0) rectangle (0.10cm,0.10cm);}}}

\newcommand{\primitive}[1]{\textsf{\textbf{#1}}}
\newcommand{\MTTM}{MTT${}_{\mathcal{M}}$}
\newcommand{\ChorMTT}{Chor${}_{\textrm{MTT}}$}
\newcommand{\ProcMTT}{Proc${}_{\textrm{MTT}}$}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Kami: A modal type theory for~distributed~systems}
\author{Maxim Urschumzew \and Miëtek Bak}
\date{\today}
\begin{document}
\maketitle
\begin{abstract}
  \noindent
  We present Kami, a programming language for distributed systems based on
  Modal Type Theory. In particular, we describe a mode theory that is suitable
  for distributed programming, and show how the primitives of Chor$\lambda$, a
  choreographic programming language, can be recovered in Kami. Finally, we
  sketch a categorical semantics that gives rise to a generic compilation
  procedure for Kami.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Modalities provide a way to extend a type theory with domain-specific type
constructors, which can be used to track runtime properties of programs
directly in the type system. For example, modalities have been used to
represent unevaluated programs for meta-programming\cite{bak2020self} and
staged computation\cite{davies2001modal}, and to encode the notion of location
in ML5, a language for distributed systems\cite{murphy2008modal}. In fact, the
``\texttt{at}'' modality of ML5 serves the same purpose as the location
annotations in choreographic programming languages\cite{giallorenzo2024choral},
and it~stands to reason that existing research on modalities could provide a
stepping stone for the design of type systems for such languages.

In this paper, we explore how Chor$\lambda$ can be based on top of a simply
typed lambda calculus with modalities, a~variant of Modal Type Theory. This
requires introducing two new concepts: common knowledge between multiple
participants, and the representation of global choreographies as values local
to individual participants.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Preliminaries}
\subsection{\texorpdfstring{Chor$\lambda$}{ChorLambda}}
Chor$\lambda$ is a functional choreographic programming language introduced by
Cruz-Filipe et~al.\cite{cruz2022functional}, and extends the simply typed
lambda calculus (STLC) with location annotations. For example, $t :
\textsf{Bool} @ r$ means that $t$ evaluates to a value of type \textsf{Bool}
located at role~$r$. The language has two communication primitives:
\primitive{com} for communicating data, and \primitive{select} for
communicating choice of branching.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Simply typed modal type theory}
Modal Type Theory (MTT), introduced by Gratzer in his PhD
thesis\cite{gratzer2023syntax}, is a framework for constructing modal type
theories, with Martin-L\"{of} type theory at its core. The~framework is
parametrized by a system of modalities specified in the form of a \emph{mode
theory}. MTT~includes so-called ``full-spectrum'' dependent types, but we
restrict ourselves to a~simply typed fragment of MTT.
\begin{definition}[{\cite[6.1.1]{gratzer2023syntax}}]
  A \emph{mode theory} is given by the following data:
  \begin{itemize}
  \item
    A set of \emph{modes} $M$. Every mode $m \in M$ instantiates a distinct
    copy of an extension of STLC.
  \item
    For each pair of modes $m, n \in M$, a set $m \to n$ of \emph{modalities}
    between the modes. A modality $\mu : m \to n$ allows us to use types and
    terms of mode $m$ in mode $n$. There may be multiple distinct modalities
    $\mu , \nu : m \to n$.
  \item
    For every pair of modalities $\mu, \nu \in m \to n$ with matching domains
    and codomains, a~set $\mu \Rightarrow \nu$ of \emph{transformations}
    between the modalities. A transformation $\alpha : \mu \Rightarrow \nu$
    allows us to convert types and terms from under the modality $\mu$ to under
    the modality $\nu$. There may be multiple distinct transformations $\alpha,
    \beta : \mu \Rightarrow \nu$.
  \end{itemize}
  Concretely, these data form a 2-category, in the sense that identity
  modalities (of type $\textsf{id}_m : m \to m$) and identity transformations
  (of type $\textsf{id}_\mu : \mu \Rightarrow \mu $) exist, modalities and
  transformations can be composed (denoted by $\mu \fatsemi \nu$ and $\alpha
  \fatsemi \beta$ respectively) if the domains and codomains match, and
  identity and associativity laws hold as expected\cite{licata2016adjoint}.
\end{definition}
\begin{definition}[{\cite[following 6.2]{gratzer2023syntax}}]
  Let $\mathcal{M}$ be a mode theory. A \emph{simply typed modal type theory}
  \MTTM{} is given by $|M|$ copies of an extension of STLC, combined as
  follows: For~each mode $m \in M$, let $\Gamma\in\textsf{Ctx}_m$, $A \in
  \textsf{Type}_m$, $t \in \textsf{Term}_m$, and $\Gamma \vdash_m t : A$ denote
  the contexts, types, terms, and typing judgements of the $m$th copy of an
  extension of STLC respectively. The rules of \MTTM{} are displayed in figures
  \ref{fig:mtt_type}, \ref{fig:mtt_ctx}, \ref{fig:mtt_var}, and
  \ref{fig:mtt_term}.
\end{definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Types
\begin{figure}
  \centering
  \begin{mathpar}
    \inferrule*[Lab=Mod-Form]
    {
      A \in \textsf{Type}_m
      \\
      \mu : m \to n
    }
    {\langle A | \mu \rangle \in \textsf{Type}_n}

    \inferrule*[Lab=Fun-Form]
    {
      A \in \textsf{Type}_m
      \\
      \mu : m \to n
      \\
      B \in \textsf{Type}_n
    }
    {(A | \mu) \to B \in \textsf{Type}_n}
    \\
    \inferrule*[Lab=Prod-Form]
    {
      A \in \textsf{Type}_m
      \\
      B \in \textsf{Type}_m
    }
    {A \times B \in \textsf{Type}_m}

    \inferrule*[Lab=Sum-Form]
    {
      A \in \textsf{Type}_m
      \\
      B \in \textsf{Type}_m
    }
    {A + B \in \textsf{Type}_m}
  \end{mathpar}
  \caption{Types of \MTTM{}}
  \label{fig:mtt_type}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Types}
Types of \MTTM{} (figure \ref{fig:mtt_type}) mirror the standard types of STLC,
with the following differences:
\begin{itemize}
\item
  A \emph{modal type} $\langle A | \mu \rangle \in \textsf{Type}_n$ is formed by
  lifting a type $A \in \textsf{Type}_m$ to mode $n$ using a modality $\mu : m
  \to n$.
\item
  A \emph{modal function type} $(A | \mu) \to B \in \textsf{Type}_n$ is formed
  for each type $A : \textsf{Type}_m$, $B : \textsf{Type}_n$, and modality $\mu
  : m \to n$. This type is a convenience that behaves as the standard function
  type composed with a modal type. We recover the standard function type using
  the identity modality $\textsf{id}_m : m \to m$.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Contexts
\begin{figure}
  \centering
  \begin{mathpar}
    \inferrule*[Lab=Ctx-Empty]
    {\textrm{ }}
    {\varepsilon \in \textsf{Ctx}_m}

    \inferrule*[Lab=Ctx-Ext]
    {
      \Gamma \in \textsf{Ctx}_n
      \\
      A \in \textsf{Type}_m
      \\
      \mu : m \to n
    }
    {\Gamma.(A | \mu) \in \textsf{Ctx}_n}

    \inferrule*[Lab=Ctx-Restr]
    {
      \Gamma \in \textsf{Ctx}_n
      \\
      \mu : m \to n
    }
    {\Gamma.\{\mu\} \in \textsf{Ctx}_m}
  \end{mathpar}
  \caption{Contexts of \MTTM{}}
  \label{fig:mtt_ctx}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Contexts}
Contexts of \MTTM{} (figure \ref{fig:mtt_ctx}) are defined in the following way:
\begin{itemize}
\item
  The empty context rule \textsc{Ctx-Empty}.
\item
  The context extension rule \textsc{Ctx-Ext} says that all types in a context
  $\Gamma \in \textsf{Ctx}_m$ must exist at a mode $m$, but they may originate
  at a different mode $n$ by way of a~modality $\mu : n \to m$. This means that
  an assumption in $\Gamma$ is of the form $(A | \mu) \in \Gamma$, where $A$ is
  a type, $n$ is an arbitrary mode, and $\mu : n \to m$ is a modality that
  brings the assumption into mode $m$. If the type originates at the current
  mode, then $\mu = \textsf{id}_m : m \to m$.
\item
  The context restriction rule \textsc{Ctx-Restr} allows us to restrict the use
  of types to those under a particular modality. That is, starting with a
  context $\Gamma \in \textsf{Ctx}_n$ and a pair of modalities $\mu, \nu : m
  \to n$, the restricted context $\Gamma.\{\nu\}\in\textsf{Ctx}_m$ only allows
  referring to the assumption $(A | \mu) \in \Gamma$ given a transformation
  $\alpha : \mu \Rightarrow \nu$. This is ensured by the variable rule
  \textsc{Var} (figure \ref{fig:mtt_term}).
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Variables
\begin{figure}
  \centering
  \begin{mathpar}
    \inferrule*[Lab=Var-Zero]
    {\ }
    {\primitive{zero} : (A | \mu) \in_{\textsf{id}_n} \Gamma.(A | \mu)}

    \inferrule*[Lab=Var-Suc]
    {
      x : (A | \mu) \in_{\nu} \Gamma
    }
    {\primitive{suc }x : (A | \mu) \in_{\nu} \Gamma.(B | \eta)}

    \inferrule*[Lab=Var-Restr]
    {x : (A | \mu) \in_{\nu} \Gamma}
    {x : (A | \mu) \in_{\eta \fatsemi \nu} \Gamma.\{\eta\}}
  \end{mathpar}
  \caption{Variables (de Bruijn indices) of \MTTM{}}
  \label{fig:mtt_var}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Terms
\begin{figure}[H]
  \centering
  \begin{mathpar}
    \inferrule*[Lab=Var]
    {
      \mu, \nu : m \to n
      \\
      \alpha : \mu \Rightarrow \nu
      \\
      x : (A | \mu) \in_\nu \Gamma
    }
    {\Gamma \vdash_n x^\alpha : A}

    \inferrule*[Lab=Mod-Intro]
    {
      \mu : m \to n
      \\
      \Gamma.\{\mu\} \vdash_m t : A
    }
    {\Gamma \vdash_n \primitive{mod}_\mu\  t : \langle A | \mu \rangle}
    \\

    \inferrule*[Lab=Mod-Elim]
    {
      \mu : m \to n
      \\
      \nu : n \to o
      \\
      \Gamma.\{\nu\} \vdash_n s : \langle A | \mu \rangle
      \\
      \Gamma.\,x : (A | \mu \fatsemi \nu) \vdash_o t : B
    }
    {
      \Gamma \vdash_o \primitive{let}_\nu\:\primitive{mod}_\mu\  x
      \bm{\leftarrow} s \primitive{ in } t : B
    }
    \\

    \inferrule*[Lab=Fun-Intro]
    {
      \mu : m \to n
      \\
      \Gamma.\,x : (A | \mu) \vdash_n t : B
    }
    {\Gamma \vdash_n \bm{\lambda\,} x \bm{\,.\,} t : (A | \mu) \to B}

    \inferrule*[Lab=Fun-Elim]
    {
      \mu : m \to n
      \\
      \Gamma \vdash_n t : (A | \mu) \to B
      \\
      \Gamma \vdash_n s : A
    }
    {\Gamma \vdash_n t\,s : B}
    \\

    \inferrule*[Lab=Prod-Intro]
    {
      \Gamma \vdash_m s : A
      \\
      \Gamma \vdash_m t : B
    }
    {\Gamma \vdash_m s \bm{\,,\,} t : A \times B}

    \inferrule*[Lab=Prod-Elim${}_1$]
    {\Gamma \vdash_m t : A \times B}
    {\Gamma \vdash_m \primitive{fst } t : A}

    \inferrule*[Lab=Prod-Elim${}_2$]
    {\Gamma \vdash_m t : A \times B}
    {\Gamma \vdash_m \primitive{snd } t : B}
    \\

    \inferrule*[Lab=Sum-Intro${}_1$]
    {\Gamma \vdash_m t : A}
    {\Gamma \vdash_m \primitive{left } t : A + B}

    \inferrule*[Lab=Sum-Intro${}_2$]
    {\Gamma \vdash_m t : B}
    {\Gamma \vdash_m \primitive{right } t : A + B}
    \\

    \inferrule*[Lab=Sum-Elim]
    {\Gamma \vdash_m s : A + B
      \\
      \Gamma.\,x : (A | \textsf{id}_m) \vdash_m t : C
      \\
      \Gamma.\,y : (B | \textsf{id}_m) \vdash_m u : C
    }
    {\Gamma \vdash_m \primitive{case } s \primitive{ of}
      \,\bm{\{}\, \primitive{left } x \bm{\,\mapsto\,} t \bm{\,;\,}
      \primitive{right } y \bm{\,\mapsto\,} u : C \,\bm{\}}
    }
  \end{mathpar}
  \caption{Terms of \MTTM{}}
  \label{fig:mtt_term}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Terms}
Terms of \MTTM{} (figure \ref{fig:mtt_term}) differ from terms of STLC as
follows:
\begin{itemize}
\item
  The modality introduction rule \textsc{Mod-Intro} ensures that terms of a
  modal type $\langle A | \mu \rangle$ may only depend on variables that are
  themselves under a~$\mu$~modality. Given a value of such type, the modality
  elimination rule \textsc{Mod-Elim} allows us to assume $(A | \mu)$. The
  \emph{framing modality} $\nu$ allows the type $A$ to exist at a different
  mode than the motive $B$\cite[6.1.9]{gratzer2023syntax}. We omit this
  modality when it is the identity.
\item
  The terms for introduction and elimination of the function type reflect that
  the type is endowed with an additional modality.
\item
  All other terms involve a single mode only, and any modalities that occur in
  the terms are the identity.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Choreographic programming with MTT}
The first, essential feature of modalities in MTT is that they restrict program
availability: When constructing a term of type $\langle A | \mu \rangle$, only
variables that are themselves under a~$\mu$~modality can be used. This leads us
naturally to the idea that we can use MTT for a distributed system with a set
of participants $\rho$, by introducing modalities $@r$ for each role $r \in
\rho$. The type $\langle A | @r \rangle$ then is interpreted as data of type
$A$, located at role $r$.

The second feature of MTT is that transformations between modalities can be
introduced in a controlled way. In order to allow a term at role $r$ to be
transformed into a term at role $s$, we must introduce a transformation
$\tau_{r,s} : @r \Rightarrow @s$ between the corresponding modalities. These
transformations can be chosen freely: we may disallow communications between
some roles, and include multiple channels between other roles.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mode theory for choreographic programming}
However, a mode theory with only $@$-modalities and transformations
$\tau_{r,s}$ is not enough to recover the full expressive power of
Chor$\lambda$. In particular, expressing the \textbf{select} operator, which is
used to notify other roles about decisions that happened locally at role $r$,
requires the following additional features:
\begin{enumerate}
\item
  We need modalities expressing the fact that some data are common knowledge
  between multiple roles. We do so by allowing arbitrary conjunctions $r_1
  \land \ldots \land r_k$ in the modality. For example, the type $\langle
  \mathbb{N} | @(r\land s) \rangle$ expresses the fact that there is a natural
  number that is known by both roles $r$ and $s$. Furthermore, a
  meet-semilattice structure arises: e.g., it holds that $r \land s \leq r$,
  since the knowledge available at $r \land s$ is a subset of the knowledge at
  $r$.
\item
  We need roles to reference data that are about to be sent to other roles. For
  this, we use a new modality $\square$, which allows roles to reference global
  choreographies locally. That is, if $A$ is a global choreography type, then
  $t : \langle A | \square \rangle$ is a local term, containing a quoted
  representation of such a choreography.
\end{enumerate}
We define our mode theory with $@$ and $\square$ modalities and common
knowledge locations as follows:
\begin{definition}
Let $\rho$ be a set of roles and $\textrm{Loc}_\rho$ be the freely generated
meet-semilattice on $\rho$. The mode theory $\mathcal{M}^\rho_{\textrm{Chor}}$
is defined as follows:
\begin{itemize}
\item
  There are two modes: $\circle$ and $\triangle$. The \emph{global mode}
  $\circle$ represents the global perspective on a choreography, encompassing
  computations and data that occur at all roles. The \emph{local mode}
  $\triangle$ represents the perspective of a single location, which may be
  a~conjunction of multiple roles, that participates in the choreography.
\item
  For each location $u \in \textrm{Loc}_\rho$, a \emph{location modality} $@u :
  \triangle \to \circle$. Each of these modalities represents a different way
  of how a local computation can be embedded in the global system. Concretely,
  $@u$ expresses that the computation exists at location $u$.
\item
  An additional \emph{quotation modality} $\square : \circle \to \triangle$,
  which allows global computations to be referenced locally.
\item
  For each location $v \in \textrm{Loc}_\rho$, a \emph{quotation
  transformation} $\textsf{quote}_v : \textsf{id}_{\triangle} \Rightarrow ({}@
  v \fatsemi \square)$, which allows a role to quote a local term to be
  evaluated by an arbitrary role $v$. This transformation happens in local mode
  $\triangle$, and its interpretation involves no communication.
\item
  For each location $u \in \textrm{Loc}_\rho$, an \emph{evaluation
  transformation} $\textsf{eval}_u : (\square \fatsemi {}@ u) \Rightarrow
  \textsf{id}_{\circle}$, which is the only one involving communication between
  roles. This transformation describes that a global choreography, available in
  quoted form at $u$, can be scheduled to be performed by all involved roles.
\item
  For each pair of locations $u, v \in \textrm{Loc}_\rho$ with $u \leq v$, a
  \emph{narrowing transformation} $\textsf{narrow}_{u,v} : @u \Rightarrow @v$.
  The intuition is that if $u$ is a sublocation of $v$, then information
  available at the former is also available at the latter.
\item
  Additional equalities governing the interactions of the transformations. For
  instance, composing $\textsf{quote}_u$ and $\textsf{eval}_u$ for the same
  role $u$ is equal to the identity transformation, since it represents a
  role communicating with itself.
\end{itemize}
\end{definition}
\begin{corollary}
  In {\upshape$\mathcal{M}^{\rho}_{\textrm{Chor}}$}, it is possible to recover
  a transformation $\tau_{u,v} : @u \Rightarrow @v$ by composing
  {\upshape$\textsf{quote}_v$} and {\upshape$\textsf{eval}_u$}. Additionally, a
  role communicating with itself is equal to the identity transformation
  {\upshape$\tau_{u,u} = \textsf{id}_{@u}$}.
\end{corollary}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Kami: An MTT-based language for choreographic programming}
Our mode theory expresses the interactions between participating roles in a
distributed system, but standard MTT instantiated with
$\mathcal{M}^{\rho}_{\textrm{Chor}}$ is not suitable to be used as a~
choreographic programming language. The problem is that MTT has no notion of
\emph{deferred transformations,} because transformations are always pushed down
the syntax tree as far as possible, and only recorded at the variables. This
means that in order to control the communication behaviour of terms, we need to
introduce a dedicated term for communications that have not been performed yet.

In our semantics, only $\textsf{eval}_v$ involves communications, so the only
rule we need to add is \textsc{Let-Eval} (figure \ref{fig:mtt_leteval}).

% Let-Eval
\begin{figure}[ht]
  \centering
  \begin{mathpar}
    \inferrule*[Lab=Let-Eval]
    {
      \nu : \circle \to m
      \\
      \Gamma.\{\nu\} \vdash_{\circle} s : \langle A | \square \fatsemi @u
      \rangle
      \\
      \Gamma.\,x : (A | \nu) \vdash_m t : B
    }
    {
      \Gamma \vdash_m \primitive{let}_\nu\:\primitive{eval}_u\ x
      \bm{\leftarrow} s \primitive{ in } t : B
    }
    \\
    \inferrule*[Lab=Let-Eval-$\beta$]
    {
      \textrm{ }
    }
    {
      \Gamma \vdash_m \primitive{let}_\nu\:\primitive{eval}_u\ x
      \bm{\leftarrow} s \primitive{ in } t \ \Longrightarrow\ \Gamma \vdash_m
      \primitive{let}_\nu\:\primitive{mod}_{(\square \fatsemi @u)}\ y
      \bm{\leftarrow} s \primitive{ in } t [ y^{\textsf{eval}_u} / x ]
    }
  \end{mathpar}
  \caption{Typing and reduction rules for deferred transformations}
  \label{fig:mtt_leteval}
\end{figure}

\begin{definition}
  Let Chor${}_{\textrm{MTT}}$ be the type theory obtained by:
  \begin{enumerate}
  \item
    Instantiating simply typed MTT with $\mathcal{M}^{\rho}_{\textrm{Chor}}$.
  \item
    Extending it with the \textsc{Let-Eval} rule.
  \item
    Restricting the reduction relation analogously to how Chor$\lambda$
    restricts the reduction relation of STLC.
  \end{enumerate}
\end{definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Informal semantics and relation to
  \texorpdfstring{Chor$\lambda$}{ChorLambda}}
In order to allow for out-of-order execution of of independent processes, we
can use the same machinery as Chor$\lambda$: we can restrict reduction rules to
ensure that communications between a pair of processes always occur in the same
order, and add an additional rewriting relation that allows for independent
processes to evaluate their terms independently.

The expressivity of Chor$\lambda$ arises from the two primitives involved in
process interaction: $\primitive{com}$ and $\primitive{select}$. Communicating
data is done using $\primitive{com}$, and can be recovered in \ChorMTT{} using
the \textsc{Let-Eval} rule in the following way:
\begin{example}
  Define a function $\textsf{com}_{A,u,v} : \langle A | @u
  \rangle \to \langle A | @v \rangle$ for each local type $A$ and pair of
  locations $u,v$.
  \begin{align*}
    \textsf{com}_{A,u,v}\ x =
    \ & \primitive{let}\:\primitive{mod}_{@u}\ y \bm{\leftarrow} x\\[-3pt]
      & \primitive{let}\:\primitive{eval}_u\ z \bm{\leftarrow}
        \primitive{mod}_{@v}\ (\primitive{mod}_{(\square \fatsemi @u)}
        \ y^{\textsf{quote}_v \,\circledast\, \textsf{id}})\\[-3pt]
      & \primitive{in } z^{\textsf{id}}
  \end{align*}
\end{example}
In Chor$\lambda$, $\primitive{select}$ is used to communicate choice of
branching. This works as follows: A process at role $r$ chooses its future
behaviour based on locally available data, and afterwards communicates its
choice using \primitive{select} statements to those processes that need to be
aware of this choice. In \ChorMTT{}, the same functionality is available, but
must be stated in reverse order. First, the required data for choosing future
behaviour are communicated from $r$ to all roles that need to be aware of this
choice. After receiving the data, all relevant processes synchronously decide
their future behaviour.
\begin{example}
  Let $A, B$ be local types, and $Z$ a global type. A function of the following
  type can be defined in \ChorMTT:
  \[
    \textsf{choice}_{A,B,Z,r} : \langle A + B | @r \rangle
    \to (\langle A | @r \rangle \to Z)
    \to (\langle B | @r \rangle \to Z)
    \to Z
  \]
\end{example}
Note that since $Z$ is global, a value $z : Z$ is a choreography potentially
involving all roles.\footnote{For the sake of brevity, our mode theory
  $\mathcal{M}^{\rho}_{\textrm{Chor}}$ as defined in this paper does not track
  which roles are involved in a given term of global type. This can be done by
  extending the mode theory with a~family of global modes.}
The semantics of $\textsf{choice}$ is: Given the knowledge of $A + B$ at role
$r$, a global behaviour $z : Z$ can be chosen for all roles, with the
additional knowledge of either the value $a : A$ or $b : B$ at role $r$. The
function can be implemented in such a way that only the information regarding
which branch is going to be chosen is communicated from $r$~to other processes,
the actual value of $a$ or $b$ stays at $r$.

The interaction of the $\square$ and $@r$ modalities are key to the definition
of \textsf{choice}. Gratzer refers to functions that allow induction ``from
under a modality'' as \emph{crisp induction principles}\cite[6.3.1]{gratzer2023syntax}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Categorical semantics}
Following Gratzer, a categorical model for MTT$_{\mathcal{M}}$ is given by a
functor $F : \mathcal{M}^{\textsf{coop}} \to \textbf{Cat}$, with additional
conditions ensuring that $F$ supports all types and terms. This definition
entails, for each modality $\mu : m \to n$, a functor $F(\mu) : F(n) \to F(m)$
modeling the \emph{contravariant} context restriction operation ${-}.\{\mu\} :
\textsf{Ctx}_n \to \textsf{Ctx}_m$. Additionally, the existence of modal types
in Gratzer's model implies the existence of a further \emph{covariant} functor
$M_\mu : F(m) \to F(n)$, such that $F(\mu)$ and $M_\mu$ are adjoint in an
appropriate sense.\footnote{In the dependently typed MTT of Gratzer, the exact
  statement is that $M_\mu$ is a \emph{dependent right adjoint}, but we can
  simplify the condition somewhat in our simply typed system.}

While such a generic model of MTT is required for Gratzer's goals, our intended
semantics, and in particular, the fact that we want to compile Kami programs
into real-world executables, leads us to consider a less general, but more
specifically useful class of models.

In the following we give the definition of our special class of models, and
sketch how MTT${}_{\mathcal{M}^\rho_{\textrm{Chor}}}$, i.e., the underlying
type theory of Kami, can be interpreted in them.
\begin{definition}
  Let $\mathcal{M}$ be a mode theory. A \emph{covariant model} for
  simply typed MTT$_{\mathcal{M}}$ is given by the following data:
  \begin{enumerate}
  \item
    A category $\mathcal{C}$ representing the compilation target.
  \item
    Closure of $\mathcal{C}$ under all type and term formers of MTT.
  \item
    A (covariant) 1-functor $G : \mathcal{M} \to \textbf{Cat}$ modeling the
    semantics of individual modes and the modal types between them.
  \item
    A family of 1-functors $\iota_m : G(m) \to \mathcal{C}$, where $m \in
    \mathcal{M}$, describing how the category $G(m)$ is represented in the
    target category.
  \item
    For each transformation $\alpha : \mu \Rightarrow \nu \in \mathcal{M}$, a
    natural transformation in the target category, of type $\tau_\alpha :
    \iota_m \circ G(\mu) \Rightarrow \iota_m \circ G(\nu)$.
  \end{enumerate}
\end{definition}
Such a covariant model is a special case of a contravariant model in the sense
of Gratzer. Specifically, we claim that:
\begin{conjecture}
  A covariant model of {\upshape MTT$_{\mathcal{M}}$} can be assembled into a
  contravariant model, by freely adjoining context restriction operators.
\end{conjecture}
For our purposes, we intend to obtain a covariant model as follows: Let
$\rho$~be a~finite set of roles, and let $\textbf{STLC}$ be the syntax category
of STLC with sum types. Viewing the set $\rho$ as a discrete category, we
denote by $\textbf{STLC}^\rho$ the functor category $\rho \to \textbf{STLC}$.
That is, an object $(\Gamma_i)_{i\in\rho}$ is a $\rho$-indexed family of
$\textbf{STLC}$ contexts, and a~morphism $(\Gamma_i)_{i\in\rho} \to
(\Delta_i)_{i\in\rho}$ is given by a $\rho$-indexed family of substitutions
$(\sigma_i : \Gamma_i \to \Delta_i)_{i\in\rho}$. The~intuition is that
$\textbf{STLC}^\rho$ describes the category of $\rho$ processes running
independently of each other, with no way to interact.

Next, we freely adjoin arrows that axiomatize synchronous communication between
the processes. To properly express these arrows, we need the following
definition:
\begin{definition}
  Let $i\in\rho$ be a role. Define
  \begin{align*}
    \delta_i({-}) : \textbf{STLC} &\to \textbf{STLC}^\rho \\
    X &\mapsto j \mapsto
    \begin{cases}
      j = i \implies &X\\
      j \neq i \implies &1
    \end{cases}
  \end{align*}
  to be the function mapping an object $X \in \textbf{STLC}$ to a family
  $X_j$, whose $i$th component is $X$, and all other components are the
  terminal object.
\end{definition}
\begin{definition} Define
  \begin{align*}
    [{-}] : \textbf{STLC}^\rho &\to \textbf{STLC} \\
    (Y_j)_{j\in\rho} &\mapsto \prod_{j \in \rho} Y_j
  \end{align*}
  to be the function mapping a family $(Y_j)_{j\in\rho} \in \textbf{STLC}^\rho$
  to the product of its components.
\end{definition}
With these definitions in hand, we can represent the type of a hypothetical
communication operation that communicates a global state $X \in
\textbf{STLC}^\rho$ from all processes to a~single process $i \in \rho$ as
follows:
\[
  \textsf{com}_{X,i} : X \to \delta_i([X])
\]
\begin{definition}
  Let the \emph{category of synchronously interacting processes}
  $\textbf{SIP}_{\rho}$ be
  defined as the category $\textbf{STLC}^\rho$ of independent processes with
  freely adjoined arrows of the shape $\textsf{com}_{X,i} : X \to
  \delta_i([X])$ for any $X \in \textbf{STLC}^\rho$ and $i \in \rho$.
\end{definition}
\begin{theorem}
  Let {\upshape$\mathcal{C}$} be a cartesian closed category. There is a
  covariant model of {\upshape MTT${}_{\mathcal{M}^\rho_{\textrm{Chor}}}$} that
  has {\upshape$\textbf{SIP}_{\rho}$} as its compilation target category.
\end{theorem}
In particular, in this model, only the transformation $\textsf{eval}_v :
(\square \fatsemi {}@ v) \Rightarrow \textsf{id}_{\circle}$ is built up from
the freely adjoined $\textsf{com}$ arrows, since it is the only one involving
communication. All other transformations are modeled by arrows already existing
as part of the cartesian closed structure of $\textbf{STLC}^\rho$.
\begin{corollary}
  There is a translation function from the syntax category of
  {\upshape MTT${}_{\mathcal{M}^\rho_{\textrm{Chor}}}$} to
  {\upshape$\textbf{SIP}_{\rho}$}.
\end{corollary}
In other words, this gives us a compilation procedure for Kami programs into
any target language with cartesian closed categorical semantics and synchronous
communication primitives.

\section*{Acknowledgements}
The Kami programming language\footnote{\url{https://nlnet.nl/project/Kami/}}
has been envisioned and initially developed together with Olivia R\"ohrig. The
authors are grateful for many hours of debate on syntax and semantics.

This project is funded through NGI Zero
Core,\footnote{\url{https://nlnet.nl/core}} a fund established by NLnet with
financial support from the European Commission's Next Generation Internet
program.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\printbibliography

\end{document}
